Machine learning topological phases in real space

### 2020-07-06 08_arxiv/main.tex
Coded simulations with fourier engineered features

The bar plots in figure \ref{feature_importances} show how informative each lattice site was in learning topological phases for experiments 1 and 2. They represent the information entropy signatures along the lattices in each SSH system. For experiment 1, only 6 lattice sites $S_1$ = (1, 2, 4, 51, 52, 54) corresponding to the two sharp peaks seen in figure \ref{feature_importances_ssh1} contributed more than 70\% of total reduction in Shannon entropy. Similarly, more than 30\% of total reduction in the Shannon entropy of eigenvector data from experiment 2 was achieved by 18 lattice sites $S_2$ = (1, 2, 3, 4, 5, 6, 47, 49, 50, 51, 52, 54, 95, 96, 97, 98, 99, 100) located at the three peaks in figure \ref{feature_importances_ssh2}.

Each of the information entropy signatures shown in figure \ref{feature_importances} captures a general pattern that persists regardless of the length of the lattice (i.e, number of unit cells) used to compute them. They are not, therefore, artifacts of particular choices of hyperparameters used to run the eigenvector-ensembling algorithm. We present the information entropy signatures for longer lattices in the section \textbf{Information entropy signatures in the macroscopic limit} in the Supplementary Material.

To see if learning the topological phases can be achieved efficiently by employing simpler models we reran experiments 1 and 2 using only the most relevant lattice sites. In our rerun of experiment 1 using only lattice sites $S_1$, mean accuracies were 97.1\% for training eigenvectors, 94.5\% for validation eigenvectors and 78.6\% for test eigenvectors. Mean accuracies obtained from eigenvector-ensembling were 98.9\% for training Hamiltonians, 98.5\% for validation Hamiltonians and 98.9\% for test Hamiltonians.

Mean accuracies for our rerun of experiment 2 using only lattice sites $S_2$ were 99.9\% for training eigenvectors, 96.0\% for validation eigenvectors and 63.7\% for test eigenvectors. Eigenvector-ensembling yields mean accuracies of 100\% for training Hamiltonians, 99.6\% for validation Hamiltonians and 90.4\% for test Hamiltonians.

These results demonstrate that learning topological phases from local real space data is still possible even for small subsets of lattice sites. We refer the reader to the section \textbf{Learning topological phases from real space data} in the Supplementary Material for a discussion of how this is possible.

@@

\end{thebibliography}

@@

%\begin{figure}
%\centering
%\subfigure[]{\label{less_feat:a}\includegraphics[width=.49\textwidth]{fig1a-6561-less-simulation_merged_winding_grid.png}}
%\subfigure[]{\label{less_feat:b}\includegraphics[width=.49\textwidth]{fig1b-6561-less-simulation_merged_winding_grid.png}}
%\caption{Probability heatmaps learned from most relevant lattice sites. (a) Probability heatmap learned in experiment 1 using only lattice %sites $S_1$.  (b) Probability heatmap learned in experiment 2 using only lattice sites $S_2$.}
%\label{less_feat}
%\end{figure}

@@

The difficulty arising from machine learning problems in high-dimensional spaces is commonly referred to as the curse of dimensionality \citesupp{bishop2006pattern}. It essentially expresses the fact that the amount of data needed to ensure a machine learning algorithm will generalize well out of its training set grows exponentially with the dimensionality of feature space.

These conflicting facets of our learning problem are harmonized by the manifold hypothesis \citesupp{cayton2005algorithms,narayanan2010sample}: even though the eigenvectors exist in a high-dimensional space (100D in the numerical experiments), they are actually much lower-dimensional surfaces (2D in the numerical experiments) embedded in this space. Furthermore, the different classes in our problem correspond to different submanifolds as can easily be seen in parameter space (this is often referred to as the manifold hypothesis for classification \citesupp{rifai2011manifold}). As we have demonstrated in the article, only a small fraction of the 2D surfaces (i.e., eigenvector lattice coordinates $v^{(j,l)}(t_1,t_2)$) were needed to retrieve the 2D parameter space phase diagrams from the 100D real space data (see figure \ref{less_feat}). In this sense, key topological information can be said to be localized on few lattice sites.

\subsection*{The eigenvector-ensembling algorithm}

We describe here in detail each step of the eigenvector-ensembling algorithm.
\vspace{.3cm}
\begin{enumerate}[topsep=0pt, partopsep=0pt]
\item[1)] \textbf{Generating Hamiltonians and winding numbers:} we start generating a number of paremeterized Hamiltonians $H(\mathbf{t})$ in real space and their corresponding winding numbers $W(\mathbf{t})$, where $\mathbf{t} = (t_1, t_2,...,t_h)$ is a vector of $h$ hopping parameters (in the simplest case of the SSH model $h$ = 2). These Hamiltonians are $N\times N$ matrices, where $N$ is twice the number of unit cells in the chain.
\item[2)] \textbf{Creating training, validation and test sets:} here we split our set of parameterized Hamiltonians and winding numbers into training, validation and test sets, as is usualy done in machine learning. More explicitly, assume our parameter $\mathbf{t}$ takes on the values $\mathbf{t}_1, \mathbf{t}_2, ..., \mathbf{t}_n$ corresponding to the Hamiltonian-winding number pairs ($H_1$, $W_1$), ..., ($H_n$, $W_n$). We partition the set \{($H_i$, $W_i$)$\mid$ $i=1,...,n$\} in three disjoint subsets: the training set, the validation set and the test set.
\item[3)] \textbf{Training on eigenvectors in real space:} since each Hamiltonian $H_i$ is represented by an $N\times N$ matrix, each one will generate $N$ eigenvectors $\mathbf{v}_i^{(1)}, \mathbf{v}_i^{(2)},...,\mathbf{v}_i^{(N)}$. Our supervised learning algorithm of choice will take as inputs the real space eigenvectors $\mathbf{v}^{(j)}_i$ of each Hamiltonian $H_i$ in the training set and be trained to learn the winding number $W_i$ of their parent Hamiltonian $H_i$. Therefore, our dataset will consist of eigenvector-winding number pairs $(\mathbf{v}_i^{(j)}, W_i)$.
\item[4)]\textbf{Eigenvector ensembling:} in order to predict the phase of a system described by a particular Hamiltonian we need to take into account how each of its eigenvectors were classified. This amounts to performing ensemble learning on the eigenvectors of each Hamiltonian. In this work we estimate the class probabilities for each Hamiltonian as the fraction of its eigenvectors that were classified in each class.
\item[5)] \textbf{Bootstrapping:} We refine the class probabilities for each Hamiltonian using a bootstrapping procedure, i.e., we repeat steps (1)-(4) $n_\text{exp}$ times, at each round sampling randomly a new training set from our grid in \textbf{t}-space. The final estimated probabilities are then arrived at by averaging the probabilities obtained in each experiment.
\end{enumerate}
\vspace{.3cm}
In this work we implemented the eigenvector ensembling algorithm with Python's Scikit tools.

@@

\begin{figure}
\centering
\subfigure[]{\label{feature_importances_ssh1_140}\includegraphics[width=.32\textwidth]{./ssh1/periodic_140_6561/plot_feature_importances.png}}
\subfigure[]{\label{feature_importances_ssh1_180}\includegraphics[width=.32\textwidth]{./ssh1/periodic_180_6561/plot_feature_importances.png}}
\subfigure[]{\label{feature_importances_ssh1_220}\includegraphics[width=.32\textwidth]{./ssh1/periodic_220_6561/plot_feature_importances.png}}
\caption{Information entropy signatures obtained for experiment 1 with higher lattice sizes. (a) Information entropy signature for 70 unit cells. (b) Information entropy signature for 90 unit cells. (c) Information entropy signature for 110 unit cells.}
\label{feature_importances_ssh1_longer_lattices}
\end{figure}
\begin{figure}
\centering
\subfigure[]{\label{feature_importances_ssh2_140}\includegraphics[width=.32\textwidth]{./ssh2/periodic_140_6561/plot_feature_importances.png}}
\subfigure[]{\label{feature_importances_ssh2_180}\includegraphics[width=.32\textwidth]{./ssh2/periodic_180_6561/plot_feature_importances.png}}
\subfigure[]{\label{feature_importances_ssh2_220}\includegraphics[width=.32\textwidth]{./ssh2/periodic_220_6561/plot_feature_importances.png}}
\caption{Information entropy signature obtained for experiment 2 with higher lattice sizes. (a) Information entropy signature for 70 unit cells. (b) Information entropy signature for 90 unit cells. (c) Information entropy signature for 110 unit cells.}
\label{feature_importances_ssh2_longer_lattices}
\end{figure}

@@

\begin{figure}
\centering
\subfigure[]{\label{cumulative_entropy_ssh2_140}\includegraphics[width=.32\textwidth]{./ssh2/periodic_140_6561/plot_cumulative_feature_importances}}
\subfigure[]{\label{cumulative_entropy_ssh2_180}\includegraphics[width=.32\textwidth]{./ssh2/periodic_180_6561/plot_cumulative_feature_importances}}
\subfigure[]{\label{cumulative_entropy_ssh2_220}\includegraphics[width=.32\textwidth]{./ssh2/periodic_220_6561/plot_cumulative_feature_importances}}
\caption{Cumulative entropy distributions distributions for higher lattice sizes in experiment 2. (a) Cumulative entropy distributions for 70 unit cells. (b) Cumulative entropy distributions for 90 unit cells. (c) Cumulative entropy distributions for 110 unit cells. Here the leaps in the cumulative entropy distributions are less pronounced than in experiment 1, but still noticeable. The three visible leaps occur at regions corresponding to the lattice sites $S_2$ in the case with 50 unit cells.}
\label{cdf2}
\end{figure}

\newpage
\begin{thebibliography}{1}
\urlstyle{rm}
\expandafter\ifx\csname url\endcsname\relax
  \def\url#1{\texttt{#1}}\fi
\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
\expandafter\ifx\csname doiprefix\endcsname\relax\def\doiprefix{DOI: }\fi
\providecommand{\bibinfo}[2]{#2}
\providecommand{\eprint}[2][]{\url{#2}}

\end{thebibliography}
\end{document}

>>>>

The bar plots in figure \ref{feature_importances} show how informative each lattice site was in learning topological phases for experiments 1 and 2. They represent the information entropy signatures along the lattices in each SSH system. For experiment 1, only six lattice sites $S_1$ = (0, 1, 3, 50, 51, 53) corresponding to the two sharp peaks seen in figure \ref{feature_importances_ssh1} contributed approximately 70\% of total reduction in Shannon entropy. Similarly, approximately 30\% of total reduction in the Shannon information entropy of eigenvector data from experiment 2 was achieved by eighteen lattice sites $S_2$ = (0, 1, 2, 3, 4, 5, 46, 48, 49, 50, 51, 53, 94, 95, 96, 97, 98, 99) distributed along the three peaks in figure \ref{feature_importances_ssh2}.

Each of the information entropy signatures shown in figure \ref{feature_importances} captures a general pattern that persists regardless of the length of the lattice (i.e, number of unit cells) used to compute them. They are not, therefore, artifacts of particular choices of hyperparameters used to run the eigenvector ensembling algorithm. We present the information entropy signatures for longer lattices in the section \textbf{Information entropy signatures in the macroscopic limit} in the Supplementary Material.

To see if learning the topological phases can be achieved efficiently by employing simpler models we reran experiments 1 and 2 using only a small subset of most relevant lattice sites. In our rerun of experiment 1 using only lattice sites $S'_1$ = (0, 50, 51, 99) (which contributed approximately 45 \% of total reduction in Shannon information entropy in experiment 1), mean accuracies were 97.0\% for training eigenvectors, 91.5\% for validation eigenvectors and 72.8\% for test eigenvectors. Mean accuracies obtained from eigenvector ensembling were 99.1\% for training Hamiltonians, 99.5\% for validation Hamiltonians and 94.5\% for test Hamiltonians.

Mean accuracies for our rerun of experiment 2 using only lattice sites $S'_2$ = (0, 1, 3, 48, 50, 51, 96, 98, 99) (which contributed approximately 20 \% of total reduction in Shannon information entropy in experiment 2) were 99.9\% for training eigenvectors, 87.7\% for validation eigenvectors and 47.3\% for test eigenvectors. Eigenvector ensembling yields mean accuracies of 100\% for training Hamiltonians, 99.5\% for validation Hamiltonians and 74.5\% for test Hamiltonians.

These results demonstrate that learning topological phases from local real space data in the bulk is still possible even for small subsets of lattice sites. We refer the reader to the section \textbf{Learning topological phases from real space data} in the Supplementary Material for a discussion of how this is possible.

@@

\end{thebibliography}
%\bibliographystyle{plain}
%\bibliographystyle{ieeetr}

@@

\begin{figure}
\centering
\subfigure[]{\label{less_feat:a}\includegraphics[width=.49\textwidth]{./ssh1/periodic_less_100_6561/merge_imshow_winding_grids_second_sim.png}}
\subfigure[]{\label{less_feat:b}\includegraphics[width=.49\textwidth]{./ssh2/periodic_less_100_6561/merge_imshow_winding_grids_sim.png}}
\caption{Phase diagrams learned from most relevant lattice sites. (a) Phase diagram learned by a combination of decision trees with eigenvector ensembling in experiment 1 using only lattice sites $S'_1$ = (0, 50, 51, 99). The accuracy achieved for Hamiltonians in the test set using only the four lattice sites in $S'_1$ was approximately 95\%.  (b) Phase diagram learned by a combination of random forests with eigenvector ensembling in experiment 2 using only lattice sites $S'_2$ = (0, 1, 3, 48, 50, 51, 96, 98, 99). The accuracy achieved for Hamiltonians in the test set using only the nine lattice sites in $S'_2$ was approximately 74\%.}
\label{less_feat}
\end{figure}

@@

The difficulty arising from machine learning problems in high-dimensional spaces is commonly referred to as the curse of dimensionality \cite{bishop2006pattern}. It essentially expresses the fact that the amount of data needed to ensure a machine learning algorithm will generalize well out of its training set grows exponentially with the dimensionality of feature space.

These apparently conflicting facets of our learning problem are harmonized by the manifold hypothesis \cite{cayton2005algorithms,narayanan2010sample}: even though the eigenvectors exist in a high-dimensional space (100D in the numerical experiments), they are actually much lower-dimensional surfaces (2D in the numerical experiments) embedded in this space. Furthermore, the different classes in our problem correspond to different submanifolds as can easily be seen in parameter space (this is often referred to as the manifold hypothesis for classification \cite{rifai2011manifold}). As we have demonstrated in the article, only a small fraction of the 2D surfaces (i.e., eigenvector lattice coordinates $v^{(j,l)}(t_1,t_2)$) were needed to retrieve the 2D parameter space phase diagrams from the 100D real space data (see figure \ref{less_feat}). In this sense, key topological information can be said to be localized on few lattice sites.

\subsection*{The eigenvector ensembling algorithm}

We describe here in detail each step of the eigenvector ensembling algorithm.
\vspace{.3cm}
\begin{enumerate}[topsep=0pt, partopsep=0pt]
\item[1)] \textbf{Generating Hamiltonians and winding numbers:} we start generating a number of paremeterized Hamiltonians $H(\mathbf{t})$ in real space and their corresponding winding numbers $W(\mathbf{t})$, where $\mathbf{t} = (t_1, t_2,...,t_h)$ is a vector of $h$ hopping parameters (in the simplest case of the SSH model $h$ = 2). These Hamiltonians are $N\times N$ matrices, where $N$ is twice the number of unit cells in the chain.
\item[2)] \textbf{Creating training, validation and test sets:} we split our set of parameterized Hamiltonians and winding numbers into training, validation and test sets, as is usualy done in machine learning. More explicitly, assume our parameter $\mathbf{t}$ takes on the values $\mathbf{t}_1, \mathbf{t}_2, ..., \mathbf{t}_n$ corresponding to the Hamiltonian-winding number pairs ($H_1$, $W_1$), ..., ($H_n$, $W_n$). We partition the set \{($H_i$, $W_i$)$\mid$ $i=1,...,n$\} in three disjoint subsets: the training set, the validation set and the test set.
\item[3)] \textbf{Training on eigenvectors in real space:} since each Hamiltonian $H_i$ is represented by an $N\times N$ matrix, each one will generate $N$ eigenvectors $\mathbf{v}_i^{(1)}, \mathbf{v}_i^{(2)},...,\mathbf{v}_i^{(N)}$. Our supervised learning algorithm of choice will take as inputs the real space eigenvectors $\mathbf{v}^{(j)}_i$ of each Hamiltonian $H_i$ in the training set and be trained to learn the winding number $W_i$ of their parent Hamiltonian $H_i$. Therefore, our dataset will consist of eigenvector-winding number pairs $(\mathbf{v}_i^{(j)}, W_i)$.
\item[4)]\textbf{Eigenvector ensembling:} in order to predict the phase of a system described by a particular Hamiltonian we need to take into account how each of its eigenvectors were classified. This amounts to performing ensemble learning on the eigenvectors of each Hamiltonian. In this work we estimate the phase probabilities for each Hamiltonian as the fraction of its eigenvectors that were classified in each phase.
\item[5)] \textbf{Bootstrapping:} We refine the phase probabilities for each Hamiltonian using a bootstrapping procedure, i.e., we repeat steps (1)-(4) $n_\text{exp}$ times, at each round sampling randomly a new training set from our grid in \textbf{t}-space. The final estimated probabilities are then arrived at by averaging the probabilities obtained in each experiment.
\end{enumerate}
\vspace{.3cm}
In this work we implemented the eigenvector ensembling algorithm with Python's Scikit tools.

@@

\begin{figure}
\centering
\subfigure[]{\label{feature_importances_ssh1_140}\includegraphics[width=.32\textwidth]{./ssh1/periodic_140_6561/plot_feature_importances.png}}
\subfigure[]{\label{feature_importances_ssh1_180}\includegraphics[width=.32\textwidth]{./ssh1/periodic_180_6561/plot_feature_importances.png}}
\subfigure[]{\label{feature_importances_ssh1_220}\includegraphics[width=.32\textwidth]{./ssh1/periodic_220_6561/plot_feature_importances.png}}
\caption{Information entropy signatures obtained for experiment 1 with higher lattice sizes. (a) Information entropy signature for 70 unit cells. (b) Information entropy signature for 90 unit cells. (c) Information entropy signature for 110 unit cells.}
\label{feature_importances_ssh1_longer_lattices}
\end{figure}
\begin{figure}
\centering
\subfigure[]{\label{feature_importances_ssh2_140}\includegraphics[width=.32\textwidth]{./ssh2/periodic_140_6561/plot_feature_importances.png}}
\subfigure[]{\label{feature_importances_ssh2_180}\includegraphics[width=.32\textwidth]{./ssh2/periodic_180_6561/plot_feature_importances.png}}
\subfigure[]{\label{feature_importances_ssh2_220}\includegraphics[width=.32\textwidth]{./ssh2/periodic_220_6561/plot_feature_importances.png}}
\caption{Information entropy signatures obtained for experiment 2 with higher lattice sizes. (a) Information entropy signature for 70 unit cells. (b) Information entropy signature for 90 unit cells. (c) Information entropy signature for 110 unit cells.}
\label{feature_importances_ssh2_longer_lattices}
\end{figure}

@@

\begin{figure}
\centering
\subfigure[]{\label{cumulative_entropy_ssh2_140}\includegraphics[width=.32\textwidth]{./ssh2/periodic_140_6561/plot_cumulative_feature_importances}}
\subfigure[]{\label{cumulative_entropy_ssh2_180}\includegraphics[width=.32\textwidth]{./ssh2/periodic_180_6561/plot_cumulative_feature_importances}}
\subfigure[]{\label{cumulative_entropy_ssh2_220}\includegraphics[width=.32\textwidth]{./ssh2/periodic_220_6561/plot_cumulative_feature_importances}}
\caption{Cumulative entropy distributions for higher lattice sizes in experiment 2. (a) Cumulative entropy distributions for 70 unit cells. (b) Cumulative entropy distributions for 90 unit cells. (c) Cumulative entropy distributions for 110 unit cells. Here the leaps in the cumulative entropy distributions are less pronounced than in experiment 1, but still noticeable. The three visible leaps occur at regions corresponding to the lattice sites $S_2$ in the case of 50 unit cells.}
\label{cdf2}
\end{figure}

%\newpage
%\bibliographystylesupp{naturemag-doi}
%\bibliographysupp{supp}

\end{document}
